Rotation effects on the Lyman-$\alpha$ line morphology in distant galaxies

### 2013-06-06 paper/Analisis.tex

\newcommand{\LCDM}{$\Lambda$CDM~}
\newcommand{\beq}{\begin{eqnarray}}  
\newcommand{\eeq}{\end{eqnarray}}  
\newcommand{\zz}{$z\sim 3$} 
\newcommand{\apj}{ApJ}  
\newcommand{\apjs}{ApJS}  
\newcommand{\apjl}{ApJL}  
\newcommand{\aj}{AJ}  
\newcommand{\mnras}{MNRAS}  
\newcommand{\mnrassub}{MNRAS accepted}  
\newcommand{\aap}{A\&A}  
\newcommand{\aaps}{A\&AS}  
\newcommand{\araa}{ARA\&A}  
\newcommand{\nat}{Nature}  
\newcommand{\physrep}{PhR}
\newcommand{\pasp}{PASP}    
\newcommand{\pasj}{PASJ}    
\newcommand{\avg}[1]{\langle{#1}\rangle}  
\newcommand{\ly}{{\ifmmode{{\rm Ly}\alpha}\else{Ly$\alpha$}\fi}}
\newcommand{\hMpc}{{\ifmmode{h^{-1}{\rm Mpc}}\else{$h^{-1}$Mpc }\fi}}  
\newcommand{\hGpc}{{\ifmmode{h^{-1}{\rm Gpc}}\else{$h^{-1}$Gpc }\fi}}  
\newcommand{\hmpc}{{\ifmmode{h^{-1}{\rm Mpc}}\else{$h^{-1}$Mpc }\fi}}  
\newcommand{\hkpc}{{\ifmmode{h^{-1}{\rm kpc}}\else{$h^{-1}$kpc }\fi}}  
\newcommand{\hMsun}{{\ifmmode{h^{-1}{\rm {M_{\odot}}}}\else{$h^{-1}{\rm{M_{\odot}}}$}\fi}}  
\newcommand{\hmsun}{{\ifmmode{h^{-1}{\rm {M_{\odot}}}}\else{$h^{-1}{\rm{M_{\odot}}}$}\fi}}  
\newcommand{\Msun}{{\ifmmode{{\rm {M_{\odot}}}}\else{${\rm{M_{\odot}}}$}\fi}}  
\newcommand{\msun}{{\ifmmode{{\rm {M_{\odot}}}}\else{${\rm{M_{\odot}}}$}\fi}}  
\newcommand{\lya}{{Lyman$\alpha$~}}
\newcommand{\clara}{{\texttt{CLARA}}~}
\newcommand{\rand}{{\ifmmode{{\mathcal{R}}}\else{${\mathcal{R}}$ }\fi}}  
\newcommand{\hs}{{\hspace{1mm}}}  
\newcommand{\kms}{\,km~s$^{-1}$}

@@

\title[Rotation in the Lyman-$\alpha$ line]{Rotation effects on the
  Lyman-$\alpha$ line morphology in distant galaxies}
\author[N. Garavito and J.E. Forero-Romero]{
\parbox[t]{\textwidth}{\raggedright 
  Nicolas Garavito-Camargo$^{1}$,
  Jaime E. Forero-Romero$^{1}$ and 
  Mark Dijkstra$^2$
}
\vspace*{6pt}\\
$^{1}$Departamento de F\'{i}sica, Universidad de los Andes, Cra. 1
No. 18A-10, Edificio Ip, Bogot\'a, Colombia \\
$^2$
}
\maketitle

@@

Due to the resonant nature of the Lyman alpha line, gas kinematics
play an important role shaping its morphology.

There is an extensive literature studying the influence of
outflow/inflow configurations in the shape of the outgoing Lyman-alpha
line.

@@

This paper is paper is structured as follows.

In this paper we experess a photon's frequency in terms of the
dimensionless variable $x\equiv (\nu -\nu_a)/\Delta\nu_\alpha$, where
$\nu_{\alpha}=2.46\times 10^{15}$ Hz is the Ly$\alpha$ resonance
frequency,  $\Delta_{\alpha} \equiv
\nu_{\alpha}\sqrt{2kT/m_pc^2}\equiv \nu_av_{\rm th} $ is the doppler
broadening of the line which depends on the neutral gas temperature
$T$ scattering the radiation or equivalently the thermal velocity
$v_{\rm th}$ of the atoms.

@@

\subsection{The implementation}
 We define a cartesian coordinate
system with its origin at the center of the sphere and the rotation
axis to be the $z$-axis, the components in the bulk velocity field, $\vec{v}
= v_{x}\hat{i} + v_{y}\hat{j} + v_{z}\hat{k}$. in the gas can be written as

\begin{subequations}
\begin{align}
    v_{x}=-\dfrac{y}{R}V_{\rm max}, \label{subeq1}\\
    v_{y}=\dfrac{x}{R}V_{\rm max}, \label{subeq2}\\
    v_{z}=0, \label{subeq3}
\end{align}
\end{subequations}

where $R$ is the radius of the sphere and $V_{\rm max}$ is the linear
velocity at the sphere's surface. The minus sign in the x-component of
the velocity indicates the direction of rotation, in this case we
assume that the angular velocity vector goes in the $\hat{k}$
direction.  The linear dependence of the velocity on the radial
distance describes the case of constant angular velocity
$\omega=V_{\rm max}/R$.

We take the polar angle $\theta$ that a unitary vector makes with the
rotation axis as defined by the dot product $\cos\theta =
{\hat{u}\cdot\hat{k}}$. In the Section \ref{sec:results} we will
present in detail how the line differs at different observing angles
$\theta$.

\subsection{Grid of Simulated Models}
\label{sec:models}

We compute the emergent Lyman-$\alpha$ line for several models with
differents values for the maximal rotational velocity, hydrogend optical
depth, dust optical depth and initial distributions of the photons
with respect to the gas. There are in total 60 models with the input
parameters summarized in Table  \ref{table:models}.

\begin{table}
\begin{center}
\begin{tabular}{ccc}\hline
Velocity (\kms) & $V_{\rm max}$&$0,\ 50,\ 100,\ 200,\ 300$\\
Hydrogen Optical Depth & $\tau_{H} $ & $10^{5},\ 10^{6},\ 10^{7}$\\
Dust Optical Depth & $\tau_{A}$ & $0$,$1$\\
Photons Distributions & & Central, Homogeneous\\
\hline
\end{tabular}
\caption{
Values for the varying input parameters in CLARA. Taking into account
all the possible combinations for these models
} 
\label{table:models}
\end{center}
\end{table}

@@

The central result of this paper is summarized in
Fig.~\ref{fig:differentvelocities} where we show that rotation has a
considerable effect on the morphology of the emergent Lyman-alpha line
both in the case where the photons are emitted at the sphere's center
and when they are initialized with an homogeneous distribution all
over the gas volume.

The results for these outgoing spectra are integrated over the whole
sphere, meaning that all the escaping photons were taking into account
regardless of the direction of the outgoing photons. Fig.~\ref{fig:differentobservers}
shows how if one gives a weight to each outgoing
photon according to its direction when escaping the gas distribution
it is possible to detect notable differences in the spectrum for
different viewing angles.

@@

\begin{figure*}
\includegraphics[width=0.45\textwidth]{Observers7t.png}
\includegraphics[width=0.45\textwidth]{Observers7tHOM.png}
\caption{Same as Figure \ref{fig:differentvelocities} but this time
  the different lines show the line morphology for different polar angles
  $\theta$ with respect to the rotation axis. In this case each photon
  has a weight dependent on their outgoing direction. All spectra correspond
  to the same $V_{\rm max}=300$\kms, Optical   Depth $\tau=10^{7}$ and
  Central Distribution without dust.
  \label{fig:differentobservers}}

\end{figure*}

In the following subsections we quantify the trends observed in
Fig. \ref{fig:differentvelocities} and
Fig. \ref{fig:differentobservers} as a function of the maximum
rotation velocity $V_{\rm max}$ and the position of the observer with
respect to the rotation angle, $\mu = \cos{\theta}$. All the results
in this section will be expressed in terms of rest frame wavelength.

@@

\begin{figure*}
     \includegraphics[width=0.40\textwidth]{2dHistogram0t5HOM.png}
     \includegraphics[width=0.40\textwidth]{2dHistogram300t5HOM.png}

    \caption{Histogram of $N_{scatt}$ vs $x$ for the homogeneous distribution, left for $V_{max}=0km/s $, right for $V_{max}=300km/s$ \label{fig:NscattHistoHOM}} 
\end{figure*}

\begin{figure*}
     \includegraphics[width=0.40\textwidth]{2dHistogram50t5.png}
     \includegraphics[width=0.40\textwidth]{2dHistogram300t5.png}

    \caption{Histogram of $N_{scatt}$ vs $x$ for the Central distribution, left for $V_{max}=50km/s $, right for $V_{max}=300km/s$ \label{fig:NscattHistoHOM}} 
\end{figure*}

>>>>

\newcommand{\LCDM}{$\Lambda$CDM~}
\newcommand{\beq}{\begin{eqnarray}}  
\newcommand{\eeq}{\end{eqnarray}}  
\newcommand{\zz}{$z\sim 3$} 
\newcommand{\apj}{ApJ}  
\newcommand{\apjs}{ApJS}  
\newcommand{\apjl}{ApJL}  
\newcommand{\aj}{AJ}  
\newcommand{\mnras}{MNRAS}  
\newcommand{\mnrassub}{MNRAS accepted}  
\newcommand{\aap}{A\&A}  
\newcommand{\aaps}{A\&AS}  
\newcommand{\araa}{ARA\&A}  
\newcommand{\nat}{Nature}  
\newcommand{\physrep}{PhR}
\newcommand{\pasp}{PASP}    
\newcommand{\pasj}{PASJ}    
\newcommand{\avg}[1]{\langle{#1}\rangle}  
\newcommand{\ly}{{\ifmmode{{\rm Ly}\alpha~}\else{Ly$\alpha$~}\fi}}
\newcommand{\hMpc}{{\ifmmode{h^{-1}{\rm Mpc}}\else{$h^{-1}$Mpc }\fi}}  
\newcommand{\hGpc}{{\ifmmode{h^{-1}{\rm Gpc}}\else{$h^{-1}$Gpc }\fi}}  
\newcommand{\hmpc}{{\ifmmode{h^{-1}{\rm Mpc}}\else{$h^{-1}$Mpc }\fi}}  
\newcommand{\hkpc}{{\ifmmode{h^{-1}{\rm kpc}}\else{$h^{-1}$kpc }\fi}}  
\newcommand{\hMsun}{{\ifmmode{h^{-1}{\rm {M_{\odot}}}}\else{$h^{-1}{\rm{M_{\odot}}}$}\fi}}  
\newcommand{\hmsun}{{\ifmmode{h^{-1}{\rm {M_{\odot}}}}\else{$h^{-1}{\rm{M_{\odot}}}$}\fi}}  
\newcommand{\Msun}{{\ifmmode{{\rm {M_{\odot}}}}\else{${\rm{M_{\odot}}}$}\fi}}  
\newcommand{\msun}{{\ifmmode{{\rm {M_{\odot}}}}\else{${\rm{M_{\odot}}}$}\fi}}  
\newcommand{\lya}{{Lyman$\alpha$~}}
\newcommand{\clara}{{\texttt{CLARA}}~}
\newcommand{\rand}{{\ifmmode{{\mathcal{R}}}\else{${\mathcal{R}}$ }\fi}}  
\newcommand{\hs}{{\hspace{1mm}}}  
\newcommand{\kms}{\,km~s$^{-1}$}

@@

\title[Rotation in the Lyman-$\alpha$ line]{Rotation effects on the
  Lyman-$\alpha$ line morphology in distant galaxies}
\author[Garavito-Camargo, Forero-Romero \& Dijkstra]{
\parbox[t]{\textwidth}{\raggedright 
  Nicolas Garavito-Camargo$^{1}$,
  Jaime E. Forero-Romero$^{1}$ and 
  Mark Dijkstra$^2$
}
\vspace*{6pt}\\
$^{1}$Departamento de F\'{i}sica, Universidad de los Andes, Cra. 1
No. 18A-10, Edificio Ip, Bogot\'a, Colombia \\
$^2$
}
\maketitle

@@

Due to the resonant nature of the Lyman alpha line, gas kinematics
play an important role shaping its morphology...

There is an extensive literature studying the influence of
outflow/inflow configurations in the shape of the outgoing Lyman-alpha
line...

@@

This paper is paper is structured as follows...

In this paper we express a photon's frequency in terms of the
dimensionless variable $x\equiv (\nu -\nu_a)/\Delta\nu_\alpha$, where
$\nu_{\alpha}=2.46\times 10^{15}$ Hz is the Ly$\alpha$ resonance
frequency,  $\Delta_{\alpha} \equiv
\nu_{\alpha}\sqrt{2kT/m_pc^2}\equiv \nu_av_{\rm th} $ is the doppler
broadening of the line which depends on the neutral gas temperature
$T$ scattering the radiation or equivalently the thermal velocity
$v_{\rm th}$ of the atoms.

@@

\subsection{Detailed Implementation of Rotation}

 In the MonteCarlo code we define a cartesian coordinate system to
 define the position of each photon. The origin of this system
 coincides withthe center of the sphere and the rotation axis is defined
 to be $z$-axis. With this choice, the components of the gas bulk velocity
 field, $\vec{v} = v_{x}\hat{i} + v_{y}\hat{j} + v_{z}\hat{k}$, can be
 written as

\begin{subequations}
\begin{align}
    v_{x}=-\dfrac{y}{R}V_{\rm max}, \label{subeq1}\\
    v_{y}=\dfrac{x}{R}V_{\rm max}, \label{subeq2}\\
    v_{z}=0, \label{subeq3}
\end{align}
\end{subequations}
%
where $R$ is the radius of the sphere and $V_{\rm max}$ is the linear
velocity at the sphere's surface. The minus/plus sign in the
$x$/$y$-component of the velocity indicates the direction of
rotation. In this case we take the angular velocity in the same
direction as the $\hat{k}$ unit vector. With these definitions we can
write the angular velocity as $\omega=V_{\rm max}/R$.

In contrast with spherically symmetric models (static, outflow,
inflow) the rotation now defines a preferred direction in the
problem. In Section \ref{sec:results} we quantify this differences by
varying the line of sight of a mock observer with respect to the
rotation axis. The results are parameterized by the polar angle
$\theta$ as defined by the dot product $\cos\theta =
{\hat{u}\cdot\hat{k}}$.

\subsection{Grid of Simulated Galaxies}
\label{sec:models}

In the Monte Carlo calculations we follow the propagation of $N_{\gamma}=10^5$
numerical photons through different spherical galaxies, each one
varying in at least one of the following parameters: the maximum
rotational velocity $V_{\rm max}$, hydrogen optical depth $\tau_{H}$,
dust optical depth $\tau_{a}$ and the initial disitrbution of photons
with respect to the gas. There are in total 60 models with the input
parameters summarized in Table \ref{table:models}.

\begin{table}
\begin{center}
\begin{tabular}{llc}\hline\hline
Physical Parameter (units) & Symbol & Values\\\hline
Velocity (\kms) & $V_{\rm max}$&$0,\ 50,\ 100,\ 200,\ 300$\\
Hydrogen Optical Depth & $\tau_{H} $ & $10^{5},\ 10^{6},\ 10^{7}$\\
Dust Optical Depth & $\tau_{a}$ & $0$,$1$\\
Photons Distributions & & Central, Homogeneous\\\hline\hline
\end{tabular}
\caption{
List of the physical parameters that define the spherical models we
have simulated using Monte Carlo calculations. For each parameter we
vary the values in the range listed in the last column. Takig into
account all the possible combinations we end up with $60$ different
models. } 
\label{table:models}
\end{center}
\end{table}

@@

The central result of this paper is summarized in
Fig.~\ref{fig:differentvelocities} that shows clearly the considerable
impact of rotation on the morphology of the emergent \ly line. Both
panels in the Figure focus on the results for $\tau_{H}=10^{7}$,
showing that the influence of rotation is present both when the
photons are either homogeneusly or centrally initialized over the gas
volume.

These results for these outgoing spectra are constructed by taking into
account all outgoing photons regardless of the direction of
propagation. In subsection \ref{subsec:direction} we report different
statistics aiming at quantifying the changes in the observed spectra
for observers with different viewing angles.

@@

In the following subsections we quantify the trends observed in
Fig. \ref{fig:differentvelocities} and
Fig. \ref{fig:differentobservers} as a function of the maximum
rotation velocity $V_{\rm max}$ and the position of the observer with
respect to the rotation angle, $\mu = \cos{\theta}$. All the results
in this section will be expressed in terms of the adimensional
frequency $x$.

@@

\begin{figure*}
     \includegraphics[width=0.40\textwidth]{2dHistogram0t5HOM.png}
     \includegraphics[width=0.40\textwidth]{2dHistogram300t5HOM.png}    
    \caption{Histogram of $N_{scatt}$ vs $x$ for the homogeneous
      distribution, left for $V_{max}=0km/s $, right for
      $V_{max}=300km/s$ \label{fig:NscattHistoHOM}}  
\end{figure*}

\begin{figure*}
     \includegraphics[width=0.40\textwidth]{2dHistogram50t5.png}
     \includegraphics[width=0.40\textwidth]{2dHistogram300t5.png}    
    \caption{Histogram of $N_{scatt}$ vs $x$ for the Central
      distribution, left for $V_{max}=50km/s $, right for
      $V_{max}=300km/s$ \label{fig:NscattHistoHOM}}  
\end{figure*}
